### 2016-08-10 doc/publications/plos16/plos16.tex

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
\marginpar{citations missing} 
This means that the large number of samples for which genome-wide association (GWA) data
is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
highlighting the need for incorporating modern compute paradigms to deal with these challenges.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype or clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}

@@

\subsection{RF}

@@

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

@@

\subsection{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algorithms `Spark ML'. These algorithms require data in a `DataFrame',
similar to DataFrames in R.

However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets.

We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we

@@

Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.

One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
size of these Vectors must also increase, and subsequently this results in higher memory requirents.

On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.

@@

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest &  &  &  \\
& RF (Spark ML) &  &  &  \\
& RF (R) & 0.456 & 2.89 & 18GB \\ \hline

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
& RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
& LR &  & 10hr 30min & 8GB\\ \hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & &  &\\
& RF(Spark ML) & &  &\\
& LR & &  &\\ \hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
& RF (Spark ML) & - & - & - \\
& LR & - & - & - \\ 
& kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is individuals x variants

@@

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

>>>>

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
\marginpar{citations missing} 
This means that the large number of samples for which genome-wide association (GWA) data
is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
highlighting the need for incorporating modern compute paradigms to deal with these challenges.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype or clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the ``wide'' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the ``curse of dimensionality''~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}

@@

\subsection{Random Forest}

@@

The benefits of random forests models are:
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

@@

\subsection{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

However, we observe that Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as
it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
Spark ML uses, which we elaborate on in the discussion.

We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and ``wide'' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes.

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

@@

Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.

One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and while RDDs
can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split the RDD into more
partitions and resource requirements will remain relatively stable. However, as the number of dimensions (variants) increases,
the size of the vectors increases. This results in higher memory-requirements as increasingly large vectors must be loaded
into memory.

On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
To build a tree, CursedForest will distribute tasks across compute nodes to calculate the information gain for every feature
in the feature-set. The result is then reduced to return the feature which gives the greatest information gain.
This process is then repeated until CursedForest has calculated every node for said tree.

@@

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
& RF - Spark ML &  &  &  \\
& RF - R & 0.456 & 2.89 & 18GB \\ \hline

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
& LR - Spark ML&  & 2hr 6min & 8GB\\ \hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
& RF - Spark ML & &  &\\
& LR - Spark ML & & 14hr 48min & 8GB \\ \hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is listed as individuals x variants.

@@

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
